Analysing the three-dimensional structure of {\em P. falciparum}

### 2017-07-23 doc/manuscript/paper.tex
more

@@

With more than 217 millions cases nearly 429 000 deaths in 2015,  malaria
remains a major disease burden in tropical and subtropical countries and an
impediment to economic development. In Africa, where more than 90\% of the
cases and deaths occurs, the effect of malaria extend far beyond direct
measures of mortality: the disease is thought to cost more than US\$21
billion a year \citep{onwejkwe:do}.

In humans, malaria is caused by a unicellular protozoan {\em Plasmodium} of
the following species: {\em P. vivax}, {\em P. malariae}, {\em P. ovale}, {\em
P. knowelis}, and with 99\% of the deaths associated to it, by far the most
deadly {\em P. falciparum}. The parasite has a complex life cycle, with
multiple stage both in the human and mosquito host. The human host is infected
with sporozoites through the bite of a infected female  {\em Anopheles}
mosquito. The sporozoites quickly migrate to the liver, where they start a two
weeks multiplication process. They are then released as merozoites in the
bloodstream, and infect red blood cells. The parasites then start their
"erythrocytic" stage, via another round of replications in red blood cells,
through an unusual process of cell division called schizogony. The parasite
first undergoes multiple rounds of nuclear replication and division  into 13
to 32 daughter cells, until the red blood cell bursts and provokes the release
of merezoites and the infective cycle starts anew. This asexual replication
cycle is responsible for the symptoms and the complications of the disease:
anemia, tertian fever, \dots

While efficacious vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citet{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinating changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms . Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at the
epigenetic and post-translational levels
\citep{cui:chromatin-mediated, duffy:role, hoeijmakers:placing,
horrocks:control, deitsch:mechanisms}.

\fixme{introduction on HiC}

We discuss below the analysis methods used and developed for studying the 3D
architecture of {\em P. falciparum}'s genome.

\section{Profiling {\em P. falciparum}'s genome architecture during the
asexual cycle}
\label{sec:data}

- Difficulty of the protocol on the AT-rich genome of p. falciparum.
- Summary of the data

\section{Analyzing contact count data}

- Normalization
- Quality assessment
- Enrichment of counts
- power-law analysis
- 
- Anything else?
    - Did we do a compartment analysis?
- Can we extend it?
  - Fit-Hic ?

\section{Inferring the 3D structures}

An important part of analysing the 3D structure of the genome of \textit{P.
falciparum} relied on having consensus models for each stage of the life
cycle. The method presented in \citet{ay:three-dimensional} relied on an
approach called ``multidimensional scaling'' (MDS), widely used in the
literature for the purpose of 3D structure inference
\citep{duan:three-dimensional, varoquaux:statistical, bau:, tanizawa:}. Each chromosomes is
modeled a series of beads, spaced ~10~kb apart. MDS-based method consists of
two steps: first convert the
contact counts $c_{ij}$ into ``wish-distances'' $\delta_{ij}$; then attempt
to place each bead in 3D such that the euclidean distances $d_{ij}$ separating
beads $i$ and $j$ matches as closely as possible the wish distance
$\delta_{ij}$.

\subsection{Constrained multi-dimensional scaling}

The method presented in \citet{ay:three-dimensional} relied on an
approach called ``multidimensional scaling'' (MDS), widely used in the
literature for the purpose of 3D structure inference
\citep{duan:three-dimensional, varoquaux:statistical, bau:, tanizawa:}. Each chromosomes is
modeled a series of beads, spaced ~10~kb apart. MDS-based method consists of
two steps: first convert the
contact counts $c_{ij}$ into ``wish-distances'' $\delta_{ij}$; then attempt
to place each bead in 3D such that the euclidean distances $d_{ij}$ separating
beads $i$ and $j$ matches as closely as possible the wish distances
$\delta_{ij}$.

\paragraph{Optimization}

Given the resulting wish-distances, we cast the inference of the consensus
structure $\mathbf{X}^{3 \times n}$ as the following constrained weighted MDS:

\begin{equation}\label{eq:mds1} 
\renewcommand{\arraystretch}{2} 
\begin{array}{ccll} 
\underset{\mathbf{X}}{\text{minimize}} & & 
\underset{(i,j) \in \mathcal{D}}{\sum} \frac{1}{\delta^2_{ij}}
\big(d_{ij}(\mathbf{X}) - \delta_{ij}\big)^2 & \\
\text{subject to} & & d_{ij} \leq r^2_\text{max} \quad \forall i, j &\,,\\
& & d_{i, i+1} \leq b_\text{max} \quad i = \{i:n|\text{chr}_i = \text{chr}_{i+1}\,,&\\ 
\end{array} 
\end{equation}

The various constraints corresponds to {\em prior} knowledge on the 3D
structure, and vary across organisms, time-points of the life cycle and
cell-type. For {\em P. falciparum}, the constraints are as follow.

\begin{itemize}
\item All beads lie in a spherical nucleus centered on the origin and of a
certain size.
\item Two adjacent beads must not be too far apart. Leveraging the work of
\citet{berger:high TOVERIFY}, we assume that $b^\text{max} = 91~nm$
\end{itemize}

\paragraph{Wish distances}

A crucial step of MDS-based method rely on finding a good set of
``wish-distances'' from Hi-C contact count data. While several approach have
been used in the litterature, most leverage bio-physics relationships either
between the genomic distnaces $s$ and the expected count, or the relationship
between the euclidean distance $d$ and the expected
count. \citet{ay:three-dimensional} observe that {\em P. falciparum}'s DNA
behave like a fractal globule polymer. \citet{grosberg:fractal} show fractal
globul polymer exhibits a well-defined relationship between its genomic length
and physical distance $d$:

$$d\sim s^{1/3}$$

The relationship between genomic distance $s$ and $c$ can be derived from the
data, by fitting a linear model between the log expected count and the genomic
distances and extracting the slope coefficient $\alpha$. We thus obtain, up to
a scaling factor $\beta$ a function to map contact counts to wish distances:

\begin{equation}
\label{eq:count-to-distance}
\delta_{ij} = \beta c^{\alpha/3}
\end{equation}

The scaling factor depends both on the sequencing depth of the libraries, but
also on the nucleus size of the organims. \citet{ay:three-dimensional} infers
3D models for a range of parameters and keeps the models that are the most
plausible, under the assumption that beads should be as uniformely distributed
in the nucleus as possible. 0-contact counts can either be ignored, or mapped
to the largest distance possible. Note that for high coverage data set, the
amount of non-interacting beads is sufficiently small that including those or
not in the optimization will make very little difference.

\subsection{Reproducing the results using a simpler model}

Two simplifications can be brought to the method proposed by
\citet{ay:three-dimensional}.

First, the ``count-to-distance'' relationship is based on two steps: (1) find
the relationship between genomic distances and counts; (2) from the
coefficient, deduce what type of polymer the DNA of this particular organism
is closest two (fractal globule or equilibrium); (3) use the relationship
between genomic distance $s$ and euclidean distance $d$ for that particular
type of polymer. Yet, it is likely that the probability of contact count
between two elements depends mostly on the average distance separating the two
beads. As noted in \citet{varoquaux:statistical}, while the relationship
between both genomic distance $s$ and expected contact count $c$ and the
relationship between genomic distance $s$ and euclidean distance $c$,
combining both yields the same count-to-distance function for both polymer
type:

\begin{equation}
\delta_{ij} = \beta c^{-1/3}
\end{equation}

While the difference may seem subtile, fitting the linear model to find
$\alpha$ is not as straightforward as it seems, as the range of genomic
distance used can have a large impact on the final coefficient.

The second simplification is to perform an unconstraint optimization. Indeed,
in practice, the constraints are fulfilled naturally by the data and thus can
be dropped from the optimization. In addition of drastically reducing the
computation time, this allows to use generic publicly available software such
as chromSDE (which implements a variant of the model defined here), or
Pastis-MDS (which implements exactly the unconstrained model defined here).

\subsection{Structure stability}

The first question that arise is simply: are structures from the same time
points but from a different start ponit more alike than structures from
different time points? To answer this question, I compute for each structure
and each time point a feature matrix, composed of a subsampled pairwise
euclidean distances of each structure. The reason for subsampling is simply so
that the resulting feature matrix fits in memory on a laptop. I then compute a
stochastic PCA on this feature matrix and plot the projection in 2D space
using the 3 first components. The results clearly show that structures from
the same time point cluster strongly. Repeating this experiment on singly
chromosomes yields the same result.

\section{3D gene set enrichment}

To assess whether groups of genes are colocolized in a 3D model,
\citet{ay:three-dimensional} leveraged a statistical method developped by
\citet{witten:XXX}, which requires labeling each pair of loci in two groups:
"close" or "far".

The authors used varying distance
thresholds (10\%, 20\% and 40\% of the nuclear diameter) to deem a locus pair
“close” and labeled all remaining
pairs in the set as “far”.

\fixme{Re-read Witten and Noble}

This approach dichotomize loci pairs into two groups, and check for the
enrichment of a label in one of the two groups. \citep{capurso:distance-based}
presents an approach that avoids this step, and instead directly estimates the
significance within the 3D model. Briefly, for a group $\mathcal{G}$, MPED
computes a test statitics:
\begin{equation*}
M = \underset{i,j \in \mathcal{G}| c_i != c_j}{\text{median}} d_{ij}\,,
\end{equation*}

\section{Volume exclusion models cannot explain }

@@

$$
\frac{1}{2} FIXME
$$

We use IMP to generate a population of structures.

\paragraph{Adding constraints on VRSM genes}

\section{Integrative analysis of gene expression and 3D structure using KernelCCA}

\citet{ay:three-dimensional}  exploited the 3D models to relate structure to
gene expression. The authors rely on a technic called ``kernelCCA'' with
attempts to find components on different view of the data in such a way that
these are coherent with one another.

- Can we extend this to other data types? Chip-seq data available from our
review? Can the results be interpretable?

\section{Discussion}
\label{sec4}

\section{Software}
\label{sec5}

Software in the form of R code, together with a sample
input data set and complete documentation is available on
request from the corresponding author (eaheron@tcd.ie).

\section{Supplementary Material}
\label{sec6}

Supplementary material is available online at
% \href{http://biostatistics.oxfordjournals.org}%
% {http://biostatistics.oxfordjournals.org}.
\url{http://biostatistics.oxfordjournals.org}.

@@

The authors thank XXX

Funding for the project was provided by BIDS.
{\it Conflict of Interest}: None declared.

>>>>

\newcommand{\OMIT}[1]{}
\newcommand{\Xb}{\textbf{X}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\Dcal}{\mathcal{D}}
\newcommand {\br}[1]{\left(#1\right)}

@@

With more than 217 millions cases nearly 429 000 deaths in 2015,  malaria
remains a major disease burden in tropical and subtropical countries and an
impediment to economic development. In Africa, where more than 90\% of the
cases and deaths occurs, the effects of malaria extend far beyond direct
measures of mortality and the disease is thought to cost more than US\$21
billion a year \citep{onwejkwe:do}.

Malaria is caused by small unicellular protozoan {\em Plasmodium}, infecting
its host through a mosquito bite. Out of the 5 species that can infect human,
{\em P. falciparum} is by the far the most deadly with 99\% of the deaths
associated to it. The parasite has a complex life cycle, with multiple stages
both in the human and mosquito host. The human host is infected with
\textit{sporozoites} through the bite of a infected female  {\em Anopheles}
mosquito. The \textit{sporozoites} quickly migrate to the liver, where they
start a two weeks multiplication process. They are then released as
\textit{merozoites} in the bloodstream and proceed to infect red blood cells. The
parasites then start their ``erythrocytic'' cycle (through \textit{rings},
\textit{trophozoites} and \textit{schizonts} stages), via another round of
replications in red blood cells, through an unusual process of cell division
called \textit{schizogony}. The parasite first undergoes multiple rounds of nuclear
replication and division into 13 to 32 daughter cells, until the red blood
cell bursts and provokes the release of \textit{merezoites} and the infective cycle
starts anew. This asexual replication cycle is responsible for the symptoms
and the complications of the disease: anemia, tertian fever, \dots

While efficacious vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citet{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinated changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms . Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at the
epigenetic and post-translational levels
\citep{cui:chromatin-mediated, duffy:role, hoeijmakers:placing,
horrocks:control, deitsch:mechanisms}.

In recent year, chromosome conformation captures-like method, broadly referred
to as Hi-C allowed to identify physical interactions between two regions in a
genome-wide fashion, yielding information on their relative spatial distance
in the nucleus. Hi-C has opened new avenues for more systematic analysis of
the three-dimensional folding of the genome, paving the way for a better
understanding of the relations between 3D structure and gene regulation,
replication timing, epigenetic changes as well as many other biological
processes. The 3D genome of a wide variety of organisms has been studied in
the past year, including many species of yeasts \citet{duan:three-dimensional,
burton:xx}, bacteria, flies, plants and numerous human and mouse cell lines.
Within the past few years, two studies of the three-dimensional structure of
{\em P. falciparum} have been released.  \citet{lemieux:genome-wide} probed
several strains of {\em P. falciparum} to understand the link between 3D
structure and the complex regulation of \textit{var} genes, a family of genes
involved the invasion of red blood cell and also responsible for the
parasite's great capacity to evade immuno-therapy.
\citet{ay:three-dimensional} assayed the 3D structure of the genome of the
parasite during three key stages of the erythrocytic cycle.
\citet{ay:multiple} reviews the key biological findings of those studies.

An important part of analysing the three-dimensional structure of the
genome of {\em P. falciparum} (as well as many other organisms) relied on the
inferring accurate 3D models of the structure. These models were then used as
a preliminary step to various analysis, such as identifying colocolized
elements, distinguishing between open and closed chromatin and last but not
least, intergrating with other sources of data, such as gene expression or
chromatin modification.

The study of chromosome organization based on contact count maps broadly
falls into two categories: model-based studies and data-driven studies. The
former methods consider the polymer nature of DNA to leverage the theoretical
and computational work done in statistical physics of polymers to build with
as few assumptions as possible many chromosome conformations. Those chromosome
conformations are then used to compare against experimental data, such as Hi-C
contact count matrices, in order to iteratively improve the models. These
models offer mechanistical insights into the folding of DNA. The latter
approaches use the experimental data to infer 3D models, by typically minizing
a cost function ensuring the models are as consistent as possible with the
data. These data driven models and analysis, with a particular focus on the
one applied to {\em P. falciparum} are the primary focus of this
review.

% XXX See rosa:computational

Though we here review some of the methods used to study and build models, this
is a very incomplete view of a blooming field. \citet{rosa:computational}
provide a more thorough (but again incomplete) overview of computational models
of genome architectures.

\section*{Three-dimensional models of DNA from Hi-C data}

In this section, we review ``data-driven'' methods to infer 3D models of DNA
from Hi-C data. These methods broadly fall into two categories: (i)
\textit{consensus} approaches, that aim at inferring a unique mean structures
best representing the contact count data; (ii) \textit{ensemble} methods that
yield a population of structures.

Both consensus and ensemble methods have benefits and drawbacks. Ensemble
approaches are more biologically accurate: Hi-C data are derived from a
population of cells, each of those with a uniquely folded 3D structure. A
population of structure thus better may represents the variability of
structures among ours cells. Yet, in addition of being more complex and
computationally intensive, ensemble approaches raise the question of
interpretability. Often, one has to fall back to interpreting the mean
structure \citep{kalhor:genome}, or a reduced set of structures
\citep{rousseau:three}. On the other hand, consensus approaches yield a single
structure recapitulating the rich information provided by Hi-C data, including
hallmarks of genome architecture shared across all cells despite the
cell-to-cell variability \citet{nagano:single-cell}. More amenable for
visualization and analysis, this average structure can be easily integrated
with other sources of data, such as RNA-seq, chip-seq etc, which are also
population-based.

After introducing some notations and common terminology, I will dive both in consensus
and population inference methods.

\subsection*{Notation}

A typical 3C/Hi-C experiment can be summarized as a contact
count matrix $\mathbf{C} \in R^{n \times n}$, where each row and column
correspond to a genomic-window (or loci), and each entry $c_{ij}$ to the
number of times loci $i$ and $j$ have been observed interacting. Contact count
matrices are subjected to biological and technical biases that need to be
corrected \citep{imakaev:iterative, cournac:normalization, yaffe:probabilistic,
hu:hicnorm}.

Methods model chromosomes as a series of beads in 3D, each bead corresponding
to a specific genomic window. We denote by $\textbf{X} \in \RR^{3 \times n}$
the coordinate matrix of the structure(s) and $x_i \in \RR^3$ corresponds to
the coordinate matrix of the i-th beads.

\subsection*{Consensus models}

Consensus methods aim at inferring a model that best represents the contact
count matrix. They model each chromosome by a chain of beads, and attempt to
place the beads in a 3D space to minimize an objective function
$\mathcal{O}(\mathbf{O}, \mathbf{C})$, sometimes
under constraints.

\begin{equation*}
\renewcommand{\arraystretch}{2}
\begin{array}{ccl}
\underset{\Xb}{\text{minimize}} & & \mathcal{O}(\mathbf{C}, \Xb)
\end{array}
\end{equation*}

\subsubsection*{Metric MDS-based methods}

Early methods \citep{dekker:capturing, duan:three-dimensional,
tanizawa:mapping, ay:three-dimensional} cast the
3D inference problem as a multidimensional scaling (MDS) problem: chromosomes
are models as a chain of beads, to place in 3D such that the distance between
bead $i$ and $j$ matches as closely as possible a ``wish-distance''
$\delta_{ij}$ derived
from contact counts.

\begin{equation*}
\renewcommand{\arraystretch}{2}
\begin{array}{ccl}
\underset{\Xb}{\text{minimize}} & & \underset{i, j \in \mathcal{D}}{\sum}
\frac{1}{w_{ij}} (d_{ij} - \delta_{ij})^2 \\
\text{subject to} & & \text{Biologically motivated constraints} \\
\end{array}
\end{equation*}

The first step is thus to find an adequate count-to-distance mapping to
convert contact counts into pairwise wish-distances $\delta_{ij}$.
\citet{dekker:capturing} models the 78 pairwise contact count of {\em S.
cerevisiae}'s chromosome III as a Worm-like chain polymer,
\citet{duan:three-dimensional} uses a linear mapping between contact counts
and wish distances, \citet{ay:three-dimensional} relies on the biophysical
properties of fractal globule polymers, \citet{tanizawa:mapping} inferred the
count-to-distance mapping by fitting the relationship using known pairwise
distances obtained through high-resolution FISH measures. While parameters of
these counts-to-distance functions vary, most of them can be summarized as a
power-law relationship : $c_{ij} = \beta d_{ij}^\alpha$.

\citet{lesne:3d} proposes a different approach to convert counts into
distances: they borrow the concept of shortest path from graph theory. A
graph is constructed from the contact map by considering each loci as a node.
Loci seen interacting are connected through an edge of weight
$\frac{1}{c_{ij}}$. The authors then compute wish-distances $\delta_{ij}$ as
the shortest path between node $i$ and $j$ in the graph. Constructing
wish-distances in such a way has two advantages: (i) low contact counts do not
contribute much to the wish-distances; (ii) the resulting wish-distances form
a distance matrix, and thus an optimal solution to the MDS can be found.

In addition of having subtle differences in the derivation of wish distances,
the different methods also vary by the inclusion of weights to reflect
confidence in ``wish-distances'' \citep{ay:three-dimensional}, or constraints
to reflect prior knowledge on the structure. Table~\ref{table:mds_detail}
summarizes the characteristic of each method.

\begin{table*}
\scriptsize
\centering
\begin{tabular}{cccccc}
\hline
\textbf{Paper} & Organism & Full or partial genome & Counts-to-distance mapping & \textbf{Constraints}
& \textbf{Weights}\\
\hline
\cite{dekker:capturing} & {\em S. cerevisiae} & Chr III & Worm-like chain behavior & - \\
\cite{duan:three-dimensional} & {\em S. cerevisiae} & Whole-genome & Linear
relationship & Adjacent beads, rDNA & - \\
\cite{tanizawa:mapping} & {\em S. pombe} & Whole-genome &
FISH-derived relationship & Adjacent beads, rDNA, telomeres, centromeres & - \\
\cite{ay:three-dimensional} & {\em P. falciparum} & Whole-genome & Fractal
globule derived relationship & adjacent
beads & $w_{ij} = \frac{1}{\delta_{ij}^2}$ \\
\cite{lesne:3d} & - & Whole-genome & \textit{ad hoc} & - & - \
\end{tabular}
\caption{Differences between MDS-based methods}{}
\label{table:mds_detail}
\end{table*}

\subsubsection*{Non-metric MDS-based method}

A crucial step of MDS-based method is the conversion of counts into
wish-distances. As described earlier, converting contact counts into
wish-distances require strong assumptions that may not be met in practice. For
example, this mapping may change from one organism to another
\citep{fudenberg:higher-order}, from one resolution to another
\citep{zhang:inference}, from one genomic distance range to another
\citep{ay:statistical}, or from one time point to another during the cell
cycle \citep{le:high-resolution, ay:three-dimensional}.

Two methods have since been proposed to alleviate this problem.
\citet{ben-elazar:spatial} proposes to cast the inference as a
\emph{non-metric MDS} \citep{kruskal:multidimensional}, where the 3D structure
is inferred jointly with the wish-distances. \fixme{except they do a bunch of
stuff to the count data}.

\citet{zhang:inference} proposes to parametrize the count-to-distance mapping
as a power-law ($d = \beta c^\alpha)$, and to infer jointly the parameters
jointly with the structure. In addition, the authors propose to add a
penalization term for non-interacting beads. The proposed penalization allows
to cast the problem as a convex optimization and thus guarantees to retrieve
the optimal structure.

\subsubsection*{Other ad-hoc optimization methods}

A series of methods cast the inference as an optimization method, where the
objective function is designed by hand to fullfill a number of properties
\citep{trieu:large, trieu:3D, trieu:MOGEN}.

For example, \citep{trieu:large} formulates an optimization problem that
attempts to place the beads such that, if a significant contact is observed
between $i$ and $j$, then $\| x_i - x_j \| < d_c$. On the other hand, if a
no significant contact is observed, then the two beads should be no closer
than $d_c$. In addition, the authors also place constraints on the minimum
distance separating two beads and maximum distances between two beads. To
satisfy all those properties, the authors propose to minimize the following
objective function:

\fixme{add damn objective function}

This method is then generalized to include \textit{trans} contact counts data
\citep{trieu:MOGEN}.

\cite{trieu:MOGEN} proposes to exploit the Lozentzian function, to formulate a
constraint-based problem. They first convert the contact counts $c_{ij}$ into
wish-distances, and cast the following optimization problem:

\begin{equation*}
\renewcommand{\arraystretch}{2}
\begin{array}{ccl}
\underset{\Xb}{\text{maximize}} & & \underset{| i - j | = 1}{\sum} \frac{c^2
c_\text{max}}{c^2 + (d_{ij} - \delta_{ij})} + 
\underset{| i - j | \neq 1}{\sum} \frac{c^2 c_{ij}}{c^2 + (d_{ij} - \delta_{ij})}
\end{array}
\end{equation*}

This objective problem attempts to minimize the difference between the
wish-distance $\delta_{ij}$ and the euclidean distance $d_{ij}$ between bead
$i$ and $j$ similarly to MDS-based method, but exploiting the property of the
Loretzian function to have very high derivative when constraints are broken.

\subsubsection*{Statistical-based models}

\citet{varoquaux:statistical} proposes a novel approach based on
statistical modeling of contact counts as random Poisson variables where the
intensity is a function of the distance: $c_{ij} \sim \text{Poisson}(\beta
d_{ij}^\alpha)$, and some perform the optimization only including a subset of
contact counts. The authors propose to cast the
inference of 3D models as maximizing the likelihood:

\begin{equation}
\renewcommand{\arraystretch}{2}
\begin{array}{cll}
\underset{\alpha, \beta, \textbf{X}}{\text{max}} &
\mathcal{L}(\mathbf{X}, \alpha, \beta) = \underset{i<j\leq n}{\sum}  c_{ij}
\alpha \log d_{ij} + c_{ij} \log \beta - \beta d_{ij}^\alpha &\\
\end{array}
\end{equation}

\subsection*{Ensemble methods}

\citet{bau:three-dimensional}
\cite{umbarger:three-dimensional}
\cite{hu:bayesian}

\subsection*{Single-cell models}

The last category of data-driven methods to infer the 3D architecture of the
genome rely on a new protocol to probe single-cell for their 3D structures
\citep{nagano:single-cell, XXX}. Single-cell Hi-C is still in its early days,
and only a handful of data sets are today publicly available. The contact maps
originating from these data sets are very sparse, and specific methods to
infer 3D structures need to be developed specifically for sc-HiC. The
challenge of constructing models from this data has been tackled very
differently by two groups.

\citet{nagano:single-cell} models the chromosome as a series of beads linked
as a flexible fiber. Loci seen interacting in the single-cell contact maps are
constrained to be in contact. The authors then sample a population of possible
structures.

\citet{paulsen:manifold} formulate a \textit{manifold based optimization},
where a low rank PSD matrix (and thus a distance matrix) is optimized to be as
close as possible to the sparse contact count matrix. Applying classical MDS
on this low rank PSD matrix then yields a 3D model of the genome.

\begin{table}[ht!]
\begin{center}
\footnotesize
\begin{tabular}{rlccccc}
\hline
\multirow{2}{*}{\emph{\footnotesize Publication}} & \multirow{2}{*}{\emph{\footnotesize Name}} &
\multirow{2}{*}{\emph{\footnotesize Con.}} &
\multirow{2}{*}{\emph{\footnotesize Pop.}}
&\multirow{2}{*}{\emph{\footnotesize MDS-based}} & \emph{\footnotesize Statistical} &
\multirow{2}{*}{\emph{\footnotesize Availability}} \\
& & & & & \emph{\footnotesize model} & \\
\hline
\hline
\footnotesize{\cite{dekker:capturing}} & & & & & & \\
\footnotesize{\cite{duan:three-dimensional}} &  & x & & x & & x \\
\footnotesize{\cite{tanizawa:mapping}} & & x & & x & & \\
\footnotesize{\cite{ay:three-dimensional}} & & x & & x & & \\
\footnotesize{\cite{ben-elazar:spatial}} & & x & & x & & x \\
\footnotesize{\cite{varoquaux:statistical}} & Pastis & x & & & x & x\\
\footnotesize{\cite{bau:three-dimensional}} & & & x & & & \\
\footnotesize{\cite{umbarger:three-dimensional}} & & x & & & &\\
\footnotesize{\cite{zhang:inference}} & chromSDE & x & & x & & x\\
\footnotesize{\cite{rousseau:three}} & & & x & & x & x\\
\footnotesize{\cite{hu:bayesian}} & Bach & x & x & & x & x\\
\footnotesize{\cite{kalhor:genome}} & & & x & & &\\
\footnotesize{\cite{lesne:3d}} & ShRec3D & x & & x & & x \\
\footnotesize{\cite{trieu:large}} & & x & & & & \\
\footnotesize{\cite{trieu:3D}} & & x & & & & \\
\footnotesize{\cite{trieu:MOGEN}} & MOGEN & x & & & & \\
\footnotesize{\cite{nagano:single-cell}} & & x & & & & \\
\footnotesize{\cite{paulsen:manifold}} & & x & & x & & x \\
\hline
\end{tabular}
\end{center}
\caption{\bf A comparison of 3D inference methods}{In this table, we summarize
properties of published methods to infer the 3D structure of the genome: (1)
is it a consensus or a population based inference? (2) Is it an MDS based
method? (3) or relies on a statistical modeling; (4) is the software
available or not (to the best of our knowledge).}
\end{table}

\section*{Modeling DNA as a flexible fibers under constraints}

In the previous section, we discussed how from a contact count map one could
infer a {\em consensus} model or a population of models that best represented
the contact map. Another set of methods approach the problem of finding 3D
models of genome architecture very differently. Instead of using data-driven
approaches, these methods model chromosomes as series of beads or flexible
fibers, and attempt to find the smallest set of constraints such that a
generated contact maps from these models match as closely as possible the
observed contact maps. As these models refine the constraints, more insight on
the hallmarks of the genome architecture are discovered.

\subsection*{A small set of constraints explain the 3D architecture of \textit{S. cerevisiae}}

The budding yeast \textit{S. cerivisiae}'s 3D structure has been extensively
studied, both through 3C-type studies \citep{dekker:capturing,
duan:three-dimensional, burton:} and through bio-imaging experiments
\citep{XX:}.

@@

\begin{equation}
\frac{1}{2} k_{\text{angle}} \sum^{N - 2}_{i = 1} \left( 1 - \frac{x_{i + 1} -
x_i}{\|x_{i + 1} - x_i\|} \cdot \frac{x_{i + 2} - x_{i + 1}}{\|x_{i + 2} -
  x_{i + 1}\|} \right)^2\,,
\end{equation}

\citet{tokuda:dynamical}

\citet{wong:how} model the budding yeast chromosomes a semi-flexible fiber
constrained in a nucleus, and applies 4 sequences specific forces on this
fiber to obtain certain properties: (1) centromeres are attached to a single
point of the nucleus by a segment; (2) telomeres are subjected to an outward
force, that pushes them towards the nuclear membrane; (3) rDNA is thicken; (4)
apply a random brownian movement. This model recovers many of the known
hallmarks of the 3D architecture of {\em S. cerevisiae}.

\subsection*{\textit{P. falciparum}'s 3D structure cannot be explained by a
constrained volume exclusion model}

Running a similar model on {\em P. falciparum}

\section*{Downstream analysis using 3D models: a highlight of the study of
{\em P. falciparum}'s 3D structure}

\subsection*{Profiling {\em P. falciparum}'s genome architecture during the
asexual cycle}
\label{sec:data}

In the past year, two groups have published novel dataset studying {\em P.
falciparum}'s 3D genome architecture using Hi-C. \citet{lemieux:genome-wide}
studied several strains associated with populations expression unique
\textit{var} genes, in order to study key folding properties of the 3D
structure relating to \textit{var} gene expression. On the other hand,
\citet{ay:three-dimensional} focus three key time-point of the development
cycle. Table~\ref{tab:data} summarizes the available data sets and their key
properties after processing the data as described in
\citet{ay:three-dimensional} and combining all available libraries for each
life-cycle stage and strain.

\subsection*{Structure stability}

The first question that arise is simply: are structures from the same time
points but from a different start ponit more alike than structures from
different time points? To answer this question, I compute for each structure
and each time point a feature matrix, composed of a subsampled pairwise
euclidean distances of each structure (so that the resulting feature matrix
fits in memory on a laptop). I then compute a stochastic PCA on this feature
matrix and plot the projection in 2D space using the 3 first components. The
results clearly show that structures from the same time point cluster
strongly. Repeating this experiment on singly chromosomes yields the same
result.

\subsection*{3D gene set enrichment}

To assess whether groups of genes are colocolized in a 3D model,
\citet{ay:three-dimensional} leveraged a statistical method developped by
\citet{witten:assessment}, which requires labeling each pair of loci in two groups:
"close" or "far".

\fixme{This is unclear. Introduce notatinos and rewrite this section}

The authors used varying distance
thresholds (10\%, 20\% and 40\% of the nuclear diameter) to deem a locus pair
“close” and labeled all remaining
pairs in the set as “far”.  The authors then compare the enrichment of loci
pairs of a group being close and far by resampling loci amongst a same
chromosome.

This approach dichotomizes loci pairs into two groups, and check for the
enrichment of a label in one of the two groups. \citep{capurso:distance-based}
presents an approach that avoids this step, and instead directly estimates the
significance within the 3D model. Briefly, for a group $\mathcal{G}$, MPED
computes a test statistics:
\begin{equation*}
M = \underset{i,j \in \mathcal{G}| c_i != c_j}{\text{median}} d_{ij}\,,
\end{equation*}

\begin{table}
\scriptsize
\centering
\begin{tabular}{cccccccc}
\hline
\textbf{Name} & \textbf{Strain} & \textbf{Stage} & \textbf{Resolution} &
\textbf{Number of contacts} & \textbf{Perc. of \textit{cis}} & \textbf{Perc of
\textit{trans}}& \textbf{Reference} \\
\hline
\hline
Ay-\textit{rings} & 3D7 & Late Rings & 10~kb & 16711552 & 43\% & 57\% &  \cite{ay:three-dimensional} \\
Ay-\textit{trophozoites} & 3D7 & Trophozoites &10~kb & 56348498 & 53\% & 47\% & \cite{ay:three-dimensional} \\
Ay-\textit{schizonts} & 3D7 & Schizonts & 10~kb & 11652832 & 55\% & 45 \% & \cite{ay:three-dimensional} \\
Lemieux-\textit{A4+} & IT/BC6+ & Rings & 25~kb & 18488252 & 19\% & 81\% & \cite{lemieux:genome-wide} \\
Lemieux-\textit{A4} & IT/3G8 & Rings &  25~kb & 19674672 & 28\% & 72\% & \cite{lemieux:genome-wide}\\
Lemieux-\textit{A44} & IT/BC6- & Rings & 25~kb & 18660594 & 25\% & 75\% & \cite{lemieux:genome-wide}\\
Lemieux-\textit{DCJ\_On} & NF54/DCJ on & Rings & 25~kb & 3098370 & 26\% & 74\% &\cite{lemieux:genome-wide} \\
Lemieux-\textit{DCJ\_Off} & NF54/DCF Off & Rings & 25~kb & 2533470 & 26\% & 73\% &\cite{lemieux:genome-wide} \\
Lemieux-\textit{B15C2} & NF54/B15C2 & Rings & 25~kb &  1022996 & 12\% & 88\% & \cite{lemieux:genome-wide}\\
\hline
\end{tabular}
\caption{Summary of the available {\em P. falciparum} Hi-C datasets}
\end{table}

\subsection*{Integrative analysis of gene expression and 3D structure using KernelCCA}

@@

{\it Conflict of Interest}: None declared.
